---
title: "Meta-analysis of murine lung regeneration studies: Slingshot pseudotime"
author: "Kent Riemondy"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: kate
    fig_caption: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```

```{r, results = 'hide', message = FALSE, warnings = FALSE}
library(ComplexHeatmap)
library(SingleCellExperiment)
library(scran)
library(scuttle)
library(scater)
library(batchelor)
library(data.table)
library(here)
library(openxlsx)
library(cowplot)
library(slingshot)
data_dir <- here("data")
formatted_data_dir <- "processed_data"
fig_dir <- "figs"
obj_dir <- "objects"
mkrs_dir <- "tables"
dir.create("objects/pt", recursive = TRUE)
source(here("results/2022-analysis/funs.R"))
```

## Pseudotime

Pseudotime analysis will be performed using [`slingshot`](https://bioconductor.org/packages/release/bioc/html/slingshot.html). The start point of the trajectories will be set to the naive AEC2 cells from the JCI insight manuscript. 

## scvi 


```{r, fig.cap="UMAP showing inferred lineages, colored by average pseudotime across all lineages"}
scvi_sce <- readRDS(file.path(obj_dir, "20221230_scvi_sce.rds"))
scvi_sce <- scvi_sce[, scvi_sce$clusters_8 != "8"]
scvi_sce$pt_labels <- ifelse(scvi_sce$cell_type == "Naive AEC2",
                             "Naive AE2",
                             scvi_sce$clusters_8)
if(!file.exists("objects/pt/scvi.rds")){
  sce.sling <- slingshot(scvi_sce,
                       start.clus = "Naive AE2",
                       clusterLabels = 'pt_labels', 
                       reducedDim = 'scvi')

  saveRDS(sce.sling, "objects/pt/scvi.rds")
} else {
  sce.sling <- readRDS("objects/pt/scvi.rds")
}

pseudo.paths <- slingPseudotime(sce.sling)

# average pseudotime over all lineages
shared.pseudo <- rowMeans(pseudo.paths, na.rm=TRUE)

embedded <- embedCurves(sce.sling, "UMAP", stretch = 3)

lin_names <- colnames(colData(sce.sling))[grepl("slingPseudotime",
                                                colnames(colData(sce.sling)))]

gg <- plotUMAP(sce.sling, colour_by=I(shared.pseudo))
#gg <- plotUMAP(sce.sling, colour_by="coarse_cell_type")
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
gg


```


```{r, fig.cap = "UMAP showing linaeages overlayed over the cell types"}
gg <- plotUMAP(sce.sling, colour_by="coarse_cell_type")
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
gg
```


```{r, fig.cap="UMAPs showing each lineage separately, colored by pseudotime for the lineage"}
plts <- list()
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  p <- plotUMAP(sce.sling, colour_by=lin) + 
    geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
  plts[[i]] <- p
}
plot_grid(plotlist = plts)
```

## scanvi


```{r, fig.cap="UMAP showing inferred lineages, colored by average pseudotime across all lineages"}

scanvi_sce <- readRDS(file.path(obj_dir, "20221230_scanvi_sce.rds"))
scanvi_sce <- scanvi_sce[, scanvi_sce$clusters_9 != "9"]
scanvi_sce$pt_labels <- ifelse(scanvi_sce$cell_type == "Naive AEC2",
                             "Naive AE2",
                             scanvi_sce$clusters_9)

if(!file.exists("objects/pt/scanvi.rds")){
  sce.sling <- slingshot(scanvi_sce,
                       start.clus = "Naive AE2",
                       clusterLabels = 'pt_labels', 
                       reducedDim = 'scanvi')


  saveRDS(sce.sling, "objects/pt/scanvi.rds")
} else {
  sce.sling <- readRDS("objects/pt/scanvi.rds")
}

pseudo.paths <- slingPseudotime(sce.sling)

# average pseudotime over all lineages
shared.pseudo <- rowMeans(pseudo.paths, na.rm=TRUE)

embedded <- embedCurves(sce.sling, "UMAP")

lin_names <- colnames(colData(sce.sling))[grepl("slingPseudotime",
                                                colnames(colData(sce.sling)))]

gg <- plotUMAP(sce.sling, colour_by=I(shared.pseudo))
gg <- plotUMAP(sce.sling, colour_by="coarse_cell_type")
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
gg
```

```{r, fig.cap = "UMAP showing linaeages overlayed over the cell types"}
gg <- plotUMAP(sce.sling, colour_by="coarse_cell_type")
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
gg
```

```{r, fig.cap="UMAPs showing each lineage separately, colored by pseudotime for the lineage"}

plts <- list()
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  p <- plotUMAP(sce.sling, colour_by=lin) + 
    geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
  plts[[i]] <- p
}
plot_grid(plotlist = plts)
```



## harmony


```{r, fig.cap="UMAP showing inferred lineages, colored by average pseudotime across all lineages"}
hm_sce <- readRDS(file.path(obj_dir, "hm_sce.rds"))
hm_sce$pt_labels <- ifelse(hm_sce$cell_type == "Naive AEC2",
                             "Naive AE2",
                             hm_sce$clusters_8)

if(!file.exists("objects/pt/harmony.rds")){
  sce.sling <- slingshot(hm_sce,
                       start.clus = "Naive AE2",
                       clusterLabels = 'pt_labels', 
                       reducedDim = 'harmony')


  saveRDS(sce.sling, "objects/pt/harmony.rds")
} else {
  sce.sling <- readRDS("objects/pt/harmony.rds")
}
pseudo.paths <- slingPseudotime(sce.sling)

# average pseudotime over all lineages
shared.pseudo <- rowMeans(pseudo.paths, na.rm=TRUE)

embedded <- embedCurves(sce.sling, "UMAP")

lin_names <- colnames(colData(sce.sling))[grepl("slingPseudotime",
                                                colnames(colData(sce.sling)))]

gg <- plotUMAP(sce.sling, colour_by=I(shared.pseudo))
gg <- plotUMAP(sce.sling, colour_by="coarse_cell_type")
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
gg

```


```{r, fig.cap = "UMAP showing linaeages overlayed over the cell types"}
gg <- plotUMAP(sce.sling, colour_by="coarse_cell_type")
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
gg
```

```{r, fig.cap="UMAPs showing each lineage separately, colored by pseudotime for the lineage"}
plts <- list()
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  p <- plotUMAP(sce.sling, colour_by=lin) + 
    geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
  plts[[i]] <- p
}
plot_grid(plotlist = plts)
```


## Summary 

Pseudotime results for scvi and scanvi integrations show trajectories that go

AEC2 -> differentiating -> AEC1   
AEC2 -> proliferating   
AEC2 -> transitional   
AEC2 -> transitional   

The harmony projection has a trajectory from AEC2 -> AEC2. 

## session info

<details><summary>Show session info</summary>

```{r code}
sessionInfo()
```

</details>

