---
title: "Meta-analysis of murine lung regeneration studies: Pub Figures"
author: "Kent Riemondy"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: kate
    fig_caption: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```

```{r, results = 'hide', message = FALSE, warning = FALSE}
library(ComplexHeatmap)
library(SingleCellExperiment)
library(scran)
library(scuttle)
library(scater)
library(data.table)
library(here)
library(cowplot)
data_dir <- here("data")
formatted_data_dir <- "processed_data"
fig_dir <- "final-figs"
dir.create(fig_dir, showWarnings = FALSE)
obj_dir <- "objects"
mkrs_dir <- "tables"
source("funs.R")
```


```{r}
sce <- readRDS(file.path(obj_dir, "20221230_scanvi_sce.rds"))
relabel_studies <- c(
  "Strunz_et_al_nc" =  "Bleomycin",
  "Riemondy_et_al_JCI-insight" = "LPS" ,
  "Kobayashi_et_al_ncb" = "Organoids"
)
sce$treatment <- factor(relabel_studies[sce$study])

id_clusters <- c(
  "3" = "AEC2",
  "8" = "AEC1",
  "2" = "BASC",
  "4" = "Proliferation",
  "5" = "Proliferation",
  "1" = "Transitional",
  "7" = "Transitional",
  "6" = "Differentiation",
  "9" = "Club"
)

sce$broad_cell_types <- factor(id_clusters[sce$clusters_9], levels = unique(id_clusters))

```

```{r}
# custom palette requested to use
rbow_pal_lst <- list(
  "red" = "#FF0000",
  "orange" = "#FF7F00",
  "green" = "#00C000",
  "yellow" = "#FFFF00",
  "light_magenta" = "#F74ED6",
  "blue_cyan" = "#007FFF",
  "dark_magenta" = "#6A1B9A",
  "cyan" = "#00FFFF",
  "blue" = "#0000FF"
#  "#FF007F",
)
rbow_pal <- unlist(rbow_pal_lst, use.names = FALSE)

bgp_pal <- c(
  "#0277BD",
  "#689F38",
  "#6A1B9A"
)
```

# Main figures 

## Figure 6A

Show plot of all cells, then cells from each study 

```{r, fig.width = 12, fig.height = 6}
trts <- unique(sce$treatment)

theme_nothing <- theme_map() + theme(legend.position = "none")

pstudy_all <- plotUMAP(sce,
                 point_size = 0.5,
                 colour_by = "treatment", 
                 order_by = "random_order") + 
  scale_color_manual(values = bgp_pal) + 
  labs(subtitle = "Combined") + 
  theme_nothing

pstudy_split <- plotUMAP(sce,
                         point_size = 0.5,
                         colour_by = "treatment",
                         other_fields = "treatment") +
  scale_color_manual(values = bgp_pal) + 
  facet_wrap(~treatment) +
  theme_nothing

p <- plot_grid(pstudy_split, pstudy_all, nrow = 1, rel_widths = c(0.75, 0.25))
save_plot(file.path(fig_dir, "6a-umap-by-study.pdf"), 
          p, 
          base_asp = 0.75, nrow = 1, ncol = 4)
p
```

## Figure 6B

Show plot of all cells, then cells from each study, labeling the clusters 

```{r, fig.width = 12, fig.height = 6}
pstudy_all <- plotUMAP(sce,
                 point_size = 0.5,
                 colour_by = "clusters_9", 
                 order_by = "random_order",
                 point_alpha = 1) + 
  scale_color_manual(values = rbow_pal) + 
  labs(subtitle = "Combined") + 
  guides(colour = guide_legend(title = NULL, override.aes = list(size = 6))) +
  theme_map()

pstudy_split <- plotUMAP(sce,
                         point_size = 0.5,
                         colour_by = "clusters_9",
                         other_fields = "treatment",
                         point_alpha = 1) +
  scale_color_manual(values = rbow_pal) + 
  facet_wrap(~treatment) +
  theme_nothing

p <- plot_grid(pstudy_split, pstudy_all, nrow = 1, rel_widths = c(0.75, 0.25))

save_plot(file.path(fig_dir, "6b-umap-by-cluster-and-study.pdf"), p,
          base_asp = 0.75, nrow = 1, ncol = 4)
p
```


## Figure 6C

Cell type markers

```{r, fig.width = 12, fig.height = 6}
plot_mrks <- function(sce, gene = "Sftpc", treatments = levels(trts), ...){
  x_rng <- range(reducedDim(sce, "UMAP")[, 1])
  y_rng <- range(reducedDim(sce, "UMAP")[, 2])
  
  plts <- lapply(seq_along(treatments), function(i) {
    tmp_sce <- suppressMessages(sce[, sce$treatment == treatments[i]])
    idx <- suppressMessages(order(logcounts(tmp_sce[gene, ])))
    brks <- suppressMessages(c(0, floor(max(logcounts(tmp_sce[gene, ])))))
    suppressMessages(plotUMAP(tmp_sce[, idx],
                              point_size = 0.5,
                              colour_by = gene,
                              other_fields = "study") +
                       scale_color_gradientn(name = NULL, colors = cloupe_cols, breaks = brks) +
                       labs(subtitle = treatments[i]) +
                       lims(x = x_rng, y = y_rng) + 
                       theme_map())
  })
  
  idx <- suppressMessages(order(logcounts(sce[gene, ])))
  brks <- suppressMessages(c(0, floor(max(logcounts(sce[gene, ])))))
  pstudy_all <- suppressMessages(plotUMAP(sce[, idx],
                                          point_size = 0.5,
                                          colour_by = gene,
                                          other_fields = "study") +
                                   scale_color_gradientn(name = NULL, 
                                                         colors = cloupe_cols,
                                                         breaks = brks) +
                                   labs(subtitle = "Combined") +
                                   theme_map())
  
  plts <- c(plts, list(pstudy_all))
  p <- plot_grid(plotlist = plts, nrow = 1)
  p
}
```

```{r, fig.width = 12, fig.height = 6}
genes_to_plot <- c(
  "Sftpc" = "aec2",
  "Pdpn" = "aec1",
  "Scgb1a1" = "basc",
  "Mki67" = "prolif",
  "Cldn4" = "transitional",
  "Sfn"= "transitional")
  
plts <- lapply(seq_along(genes_to_plot)[1], function(i) {
  ct = genes_to_plot[i]
  gn = names(genes_to_plot)[i]
  p <- plot_mrks(sce, gn)
  
  save_plot(
    file.path(fig_dir, paste0("6c-", ct, "-", gn, ".pdf")),
    p,
    base_asp = 1,
    nrow = 1,
    ncol = 4
  )
  p
})

plts
```


## Figure 6E

Not sure how best to show Igfbp2 expression, as it is more diffuse in this UMAP compared to previous analysis. 

```{r, fig.width = 12, fig.height = 6}
gene <- "Igfbp2"
p <- plot_mrks(sce, gene)
save_plot(
  file.path(fig_dir, paste0("6e-", gene[1], ".pdf")),
  p,
  base_asp = 1,
  nrow = 1,
  ncol = 4
)
p
```


## Figure 6F

Not sure how best to show AEC1 population. For now just replicating 6B.

```{r, fig.width = 12, fig.height = 6}
pstudy_all <- plotUMAP(sce,
                 point_size = 0.5,
                 colour_by = "treatment", 
                 order_by = "random_order") + 
  scale_color_manual(values = bgp_pal) + 
  labs(subtitle = "Combined") + 
  theme_nothing

pstudy_split <- plotUMAP(sce,
                         point_size = 0.5,
                         colour_by = "treatment",
                         other_fields = "treatment") +
  scale_color_manual(values = bgp_pal) + 
  facet_wrap(~treatment) +
  theme_nothing

p <- plot_grid(pstudy_split, pstudy_all, nrow = 1, rel_widths = c(0.75, 0.25))
save_plot(file.path(fig_dir, "6f-umap-by-study-aec-all.pdf"), p, base_asp = 0.75, nrow = 1, ncol = 4)
p
```

```{r, fig.width = 12, fig.height = 6}
zoom_x <- c(11.25, 11.75)
zoom_y <- c(1.25, 1.75)
u <- reducedDims(sce)$UMAP
pass_x <- u[, 1] > zoom_x[1] & u[, 1] < zoom_x[2]
pass_y <- u[, 2] > zoom_y[1] & u[, 2] < zoom_y[2]
mx <- mean(u[pass_x & pass_y, 1])
my <- mean(u[pass_x & pass_y, 2])
sf <- 0.05
pstudy_all <- plotUMAP(sce,
                       colour_by = "treatment", 
                       order_by = "random_order", 
                       point_size = 3) + 
  scale_color_manual(values = bgp_pal) + 
  labs(subtitle = "Combined") + 
  lims(x = c(mx - sf, mx + sf), 
       y = c(my - sf, my + sf)) + 
  theme_nothing 

pstudy_split <- plotUMAP(sce,
                         point_size = 3,
                         colour_by = "treatment",
                         other_fields = "treatment") +
  lims(x = c(mx - sf, mx + sf), 
       y = c(my - sf, my + sf)) + 
  scale_color_manual(values = bgp_pal) + 
  facet_wrap(~treatment) +
  theme_nothing

p <- plot_grid(pstudy_split, pstudy_all, nrow = 1, rel_widths = c(0.75, 0.25))
save_plot(file.path(fig_dir, "6f-umap-by-study-aec-zoom.pdf"), p, base_asp = 0.75, nrow = 1, ncol = 4)
p
```

## 6H

```{r, fig.width = 12, fig.height = 6}
x_rng <- range(reducedDim(sce, "UMAP")[, 1])
y_rng <- range(reducedDim(sce, "UMAP")[, 2])
  
ctypes <- levels(sce$broad_cell_types)
ctypes <- setdiff(ctypes, "Club")
plts <- lapply(seq_along(ctypes), function(i) {
  tmp_sce <- sce
  ct <- ctypes[i]
  tmp_sce$tmp_cat <- ifelse(tmp_sce$broad_cell_types == ct, ct, "zzz") 
  tmp_sce$tmp_cat <- factor(tmp_sce$tmp_cat, levels = c("zzz", ct))
  suppressMessages(plotUMAP(tmp_sce[, order(tmp_sce$tmp_cat)],
                            point_size = 0.5,
                            colour_by = "tmp_cat") +
                     scale_color_manual(values = c("grey50", 
                                                   sccol_pals$tableau10medium[[1]])) +
                     labs(subtitle = ct) +
                     lims(x = x_rng, y = y_rng) + 
                     theme_nothing)
  })
  
p <- plot_grid(plotlist = plts, nrow = 2, ncol = 3)
save_plot(file.path(fig_dir, "6h-umap-by-broad-cell-type.pdf"), 
          p, base_asp = 1, base_height = 3, nrow = 2, ncol = 3)
p
```

## 7A

```{r}
pstudy_all <- plotUMAP(sce,
                 point_size = 0.5,
                 colour_by = "clusters_9", 
                 order_by = "random_order") + 
  labs(subtitle = "Combined") + 
  theme_nothing

pc17 <- plotUMAP(sce[, sce$clusters_9 %in% c("1", "7")],
                         point_size = 0.5,
                         colour_by = "clusters_9",
                         other_fields = "treatment") + 
         scale_color_manual(values = rbow_pal[c(1, 7)]) + 
  guides(colour = guide_legend(title = NULL, override.aes = list(size = 6))) +
  labs(subtitle = "Transitional clusters") + 
  lims(x = x_rng, y = y_rng) + 
  theme_map()

p <- plot_grid(pstudy_all, pc17, nrow = 1)
save_plot(file.path(fig_dir, "7a-subclusters.pdf"), p, base_asp = 0.75, nrow = 1, ncol = 2)
p
```

## 7B

```{r}
fig_7b_genes <- c("Cdkn1a", "Cdkn2a", "Cdkn2b", "Trp53")

fig_7d_genes <- c("Fn1", "Tmsb10", "Clu", "Ctgf", "Igfbp7", "Itgb6", "Sox4", "Crlf1", "Pmepa1", "Adgrg1", "Tmsb4x", "Sfn", "Lgals1", "Rab1a", "Cldn4", "Rsrp1", "Sparc", "Fblim1", "Cyba", "Ramp1", "Pdgfb", "Serpinb9", "Lpcat4", "Ntm", "Marcksl1", "Mcam", "Pdlim7")

fig_7j_genes <- c("Krt17", "Tpm1", "Sox4", "Cald1", "Sfn", "Fhl2", "Plau", "Phlda2", "Itga2", "Lamb3", "S100a2", "Atf3", "Zfp36l2", "Tpm4", "Palld", "Cdh3", "Ier3")

 fig_7l_genes <- c("Hopx", "Igfbp2", "Pdpn", "Sftpc", "Lamp3", "Sftpb", "Sftpd", "Abca3", "Itgb6", "Cldn4", "Krt8", "Krt7", "Tpm1", "Cald1", "Fhl2", "S100a2", "Tpm4", "Krt17", "Sox4", "Lamb3", "Zfp36l2", "Itga2")



genes_to_plot <- list(fig_7b_genes = fig_7b_genes,
                      fig_7d_genes = fig_7d_genes,
                      fig_7j_genes = fig_7j_genes,
                      fig_7l_genes = fig_7l_genes)

genes_to_plot <- lapply(genes_to_plot, function(x){
  missing_genes <- setdiff(x, rownames(sce))
  if(length(missing_genes) > 0){
    msg <- paste0(missing_genes, collapse = ", ")
    warning("unable to find ", msg, " in matrix")
  }
  intersect(x, rownames(sce))
})


get_mean_zvals <- function(sce, genes, group = "clusters_9", clip = 2.5){
  zs <- t(scale(t(as.matrix(logcounts(sce[genes, ])))))
  avg_zs <-  sapply(split(colnames(zs), sce[[group]]), function(i) rowMeans(zs[, i]))
  avg_zs[avg_zs > clip] <- clip
  avg_zs[avg_zs < -clip] <- -clip
  avg_zs
}


gns <- genes_to_plot$fig_7b_genes
zs <- get_mean_zvals(sce, gns)
Heatmap(zs[, c(7, 1)], 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```

```{r}
Heatmap(zs, 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```



## 7D

```{r}
gns <- genes_to_plot$fig_7d_genes
zs <- get_mean_zvals(sce, gns)
to_highlight <- ifelse(gns %in% c("Fblim1", "Pdgfb", "Mcam", "Pdlim7"),
                       "red", "black")

Heatmap(zs[, c(7, 1)], 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        row_names_gp = gpar(col = to_highlight))
```

```{r}
Heatmap(zs, 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE, 
        row_names_gp = gpar(col = to_highlight))
```


## 7J

```{r}
gns <- genes_to_plot$fig_7j_genes
zs <- get_mean_zvals(sce, gns)

Heatmap(zs[, c(7, 1)], 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```

```{r}
Heatmap(zs, 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```


## 7l

```{r}
gns <- genes_to_plot$fig_7l_genes
zs <- get_mean_zvals(sce, gns)

Heatmap(zs[, c(3, 1, 7)], 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```


```{r}
Heatmap(zs, 
        name = "Mean\nZ-score", 
        col = viridis::viridis(101),
        cluster_rows = FALSE,
        cluster_columns = FALSE)
```


## 7C

```{r}
library(monocle3)
cds <- readRDS("objects/pt/monocle_scanvi.rds")
p <- plot_cells(cds, 
           color_cells_by = "clusters_9",
           label_cell_groups = FALSE,
           label_leaves = TRUE, 
           label_branch_points = FALSE,
           alpha = 0.66) +
  scale_color_manual(values = rbow_pal) +
  theme_nothing

p
save_plot(file.path(fig_dir, "7c-monocle.pdf"), p, base_asp = 1)
```

```{r slingshot}
library(slingshot)
sce.sling <- readRDS("objects/pt/scanvi.rds")
embedded <- embedCurves(sce.sling, "UMAP")
lin_names <- colnames(colData(sce.sling))[grepl("slingPseudotime",
                                                colnames(colData(sce.sling)))]
```

```{r}
gg <- plotUMAP(sce.sling, colour_by="clusters_9", point_size = 0.5)
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  gg <- gg + geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1)
}
p <- gg + theme_nothing
p
save_plot(file.path(fig_dir, "7c-slingshot.pdf"), p, base_asp = 1)
```

## additional

```{r}
plts <- list()
for(i in seq_along(lin_names)){
  lin <- lin_names[i]
  scurve <- slingCurves(embedded)[[i]] # only 1 path.
  curve_coords <- data.frame(scurve$s[scurve$ord,])
  p <- plotUMAP(sce.sling, colour_by=lin) + 
    geom_path(data=curve_coords, aes(x=Dim.1, y=Dim.2), linewidth=1) +
    scale_color_viridis_c(name = paste0("Path ", i)) + 
    theme_map()
  plts[[i]] <- p
}
p <- plot_grid(plotlist = plts, nrow = 2, ncol = 2)

save_plot(
    file.path(fig_dir, "additional-slingshot.pdf"),
    p,
    base_asp = 1.2,
    nrow = 2,
    ncol = 2
  )
p

```


```{r, fig.width = 12, fig.height = 6}
x_rng <- range(reducedDim(sce, "UMAP")[, 1])
y_rng <- range(reducedDim(sce, "UMAP")[, 2])
  
clusts <- c("7", "1")
plts <- lapply(seq_along(clusts), function(i) {
  tmp_sce <- sce
  ct <- clusts[i]
  tmp_sce$tmp_cat <- ifelse(tmp_sce$clusters_9 == ct, ct, "zzz") 
  tmp_sce$tmp_cat <- factor(tmp_sce$tmp_cat, levels = c("zzz", ct))
  suppressMessages(plotUMAP(tmp_sce[, order(tmp_sce$tmp_cat)],
                            point_size = 0.5,
                            colour_by = "tmp_cat") +
                     scale_color_manual(values = c("grey50", 
                                                   sccol_pals$tableau10medium[[1]])) +
                     labs(subtitle = paste0("cluster-", ct)) +
                     lims(x = x_rng, y = y_rng) + 
                     theme_nothing)
  })
  
p <- plot_grid(plotlist = plts, nrow = 1, ncol = 2)
save_plot(file.path(fig_dir, "additional-umap-by-cluster7_and_1.pdf"), 
          p, base_asp = 1, base_height = 3, nrow = 1, ncol = 2)
p
```

# Supplemental figures

## 6AB

```{r, fig.width = 12, fig.height = 6}
genes_to_plot <- c(
  "Lyz2" = "aec2",
  
  "Clic5" = "aec1",
  "Igfbp2" = "aec1",
  "Hopx" = "aec1",
  "Emp2" = "aec1",
  "Akap5" = "aec1",
  "Ager" = "aec1",
  
  "Scgb3a2" = "basc",
  
  "Top2a" = "prolif",
  "Krt8" = "transitional")
  
plts <- lapply(seq_along(genes_to_plot), function(i) {
  ct = genes_to_plot[i]
  gn = names(genes_to_plot)[i]
  p <- plot_mrks(sce, gn)
  
  save_plot(
    file.path(fig_dir, paste0("supplement-6ab-", ct, "-", gn, ".pdf")),
    p,
    base_asp = 0.75,
    nrow = 1,
    ncol = 4
  )
  p
})

plts
```

## 6C

```{r}
strunz_mdat <- colData(readRDS(file.path(data_dir,
                                         "Strunz_et_al_nc", 
                                         "sce.rds")))
sce$strunz_et_al_timepoint <- strunz_mdat[colnames(sce), ]$time_point
```

### bleo

```{r}
study <- "Strunz_et_al_nc"
study_cell_types <- unique(sce[, sce$study == study]$cell_type)
plts <- list()
for(i in seq_along(study_cell_types)){
  ct <- study_cell_types[i]
  tmp_sce <- sce
  tmp_sce$plt <- NA
  to_plot <- tmp_sce$cell_type == ct & tmp_sce$study == study
  tmp_sce$plt[to_plot] <- TRUE
  tmp_sce <- tmp_sce[, order(tmp_sce$plt, na.last = FALSE)]
  p <- plotUMAP(tmp_sce, 
                colour_by = "plt",
                point_size = 0.5) +
    scale_color_manual(values = sccol_pals$tableau10medium[[1]]) + 
    labs(subtitle = ct) + 
    theme_nothing +
    theme(plot.subtitle = element_text(size = 10)) 
    
  plts[[i]] <- p
}

tmp_sce <- sce
tmp_sce$plt <- NA
to_plot <- which(tmp_sce$strunz_et_al_timepoint == "d14_PBS" & 
  tmp_sce$study == study)

tmp_sce$plt[to_plot] <- TRUE
tmp_sce <- tmp_sce[, order(tmp_sce$plt, na.last = FALSE)]
p <- plotUMAP(tmp_sce,
              colour_by = "plt",
              point_size = 0.5) +
  scale_color_manual(values = viridis::viridis(10)[3]) + 
  labs(subtitle = "PBS") + 
  theme_nothing +
  theme(plot.subtitle = element_text(size = 10)) 
  
plts <- append(plts, list(p))
nc <- 3
nr <- ceiling((length(study_cell_types) + 1) / nc)


p <- plot_grid(plotlist = plts, 
               nrow = nr,
               ncol = nc)
save_plot(
    file.path(fig_dir, "supplement-6c-bleo-celltypes.pdf"),
    p,
    base_asp = 0.75,
    nrow = nr,
    ncol = nc
  )
p
```

### lps

```{r}
study <- "Riemondy_et_al_JCI-insight"
study_cell_types <- unique(sce[, sce$study == study]$cell_type)
plts <- list()
for(i in seq_along(study_cell_types)){
  ct <- study_cell_types[i]
  tmp_sce <- sce
  tmp_sce$plt <- NA
  to_plot <- tmp_sce$cell_type == ct & tmp_sce$study == study
  tmp_sce$plt[to_plot] <- TRUE
  tmp_sce <- tmp_sce[, order(tmp_sce$plt, na.last = FALSE)]
  p <- plotUMAP(tmp_sce, 
                colour_by = "plt",
                point_size = 0.5) +
    scale_color_manual(values = viridis::viridis(10)[3]) + 
    labs(subtitle = ct) + 
    theme_nothing +
    theme(plot.subtitle = element_text(size = 10)) 
    
  plts[[i]] <- p
}

tmp_sce <- sce
tmp_sce$plt <- NA
to_plot <- tmp_sce$cell_type %in% c("Other Injured AEC2",
                                     "Naive AEC2") & 
  tmp_sce$study == study

tmp_sce$plt[to_plot] <- TRUE
tmp_sce <- tmp_sce[, order(tmp_sce$plt, na.last = FALSE)]
p <- plotUMAP(tmp_sce,
              colour_by = "plt",
              point_size = 0.5) +
  scale_color_manual(values = viridis::viridis(10)[3]) + 
  labs(subtitle = "Other injured AEC2 + Naive AEC2") + 
  theme_nothing +
  theme(plot.subtitle = element_text(size = 10)) 
  
plts <- append(plts, list(p))
nc <- 3
nr <- ceiling((length(study_cell_types) + 1) / nc)


p <- plot_grid(plotlist = plts, 
               nrow = nr,
               ncol = nc)
save_plot(
    file.path(fig_dir, "supplement-6c-lps-celltypes.pdf"),
    p,
    base_asp = 0.75,
    nrow = nr,
    ncol = nc
  )
p
```

```{r}
study <- "Kobayashi_et_al_ncb"
study_cell_types <- unique(sce[, sce$study == study]$cell_type)
plts <- list()
for(i in seq_along(study_cell_types)){
  ct <- study_cell_types[i]
  tmp_sce <- sce
  tmp_sce$plt <- NA
  to_plot <- tmp_sce$cell_type == ct & tmp_sce$study == study
  tmp_sce$plt[to_plot] <- TRUE
  tmp_sce <- tmp_sce[, order(tmp_sce$plt, na.last = FALSE)]
  p <- plotUMAP(tmp_sce, 
                colour_by = "plt",
                point_size = 0.5) +
    scale_color_manual(values = viridis::viridis(10)[3]) + 
    labs(subtitle = ct) + 
    theme_nothing +
    theme(plot.subtitle = element_text(size = 10)) 
    
  plts[[i]] <- p
}

nc <- 3
nr <- ceiling(length(study_cell_types) / nc)


p <- plot_grid(plotlist = plts, 
               nrow = nr,
               ncol = nc)
save_plot(
    file.path(fig_dir, "supplement-6c-organoid-celltypes.pdf"),
    p,
    base_asp = 0.75,
    nrow = nr,
    ncol = nc
  )
p
```

```{r}
pal <- colorRampPalette(scales::brewer_pal(pal = "Spectral")(11))
p <- plotUMAP(sce,
              point_size = 0.5,
              colour_by = "cell_type",
              other_fields = "treatment", 
              point_alpha = 1) +
  facet_wrap(~treatment) +
  scale_color_manual(values = pal(length(unique(sce$cell_type)))) + 
  guides(colour = guide_legend(title = NULL, override.aes = list(size = 2))) +
  theme_map(font_size = 8)

save_plot(
    file.path(fig_dir, "supplement-6c-all-celltypes.pdf"),
    p,
    base_asp = 0.75,
    nrow = 1,
    ncol = 4
  )
p

```

## 7C

```{r}
to_plot <- c("1", "7")
p <- plotUMAP(sce[, sce$clusters_9 %in% to_plot],
              point_size = 0.5,
              colour_by = "clusters_9",
              other_fields = "treatment", 
              point_alpha = 1) +
  facet_wrap(~treatment) +
  scale_color_manual(values = rbow_pal[c(1, 7)]) + 
  guides(colour = guide_legend(title = NULL, override.aes = list(size = 2))) +
  theme_map(font_size = 8)

save_plot(
    file.path(fig_dir, "supplement-7c-transitional-celltypes.pdf"),
    p,
    base_asp = 0.75,
    nrow = 1,
    ncol = 3
  )
p
```

## 7D

```{r}
to_plot <- c("Krt19", "Ndrg1")
plts <- lapply(to_plot, function(x) {
  idx <- suppressMessages(order(logcounts(sce[x, ])))
  plotUMAP(sce[, idx],
         point_size = 0.5,
         colour_by = x) +
  scale_color_gradientn(name = x, colors = cloupe_cols) +
    theme_map()
})
p <- plot_grid(plotlist = plts, nrow = 1, ncol = 2)
save_plot(
    file.path(fig_dir, "supplement-7d.pdf"),
    p,
    base_asp = 1.2,
    nrow = 1,
    ncol = 2
  )
p
```

## 7E 

```{r}
library(monocle3)

x_rng <- range(reducedDim(sce, "UMAP")[, 1])
y_rng <- range(reducedDim(sce, "UMAP")[, 2])
  
cds <- readRDS("objects/pt/monocle_scanvi.rds")

cds$treatment <- factor(relabel_studies[colData(cds)$study])

plts <- lapply(levels(cds$treatment), function(x) {
  plot_cells(cds[, colData(cds)$treatment == x], 
           color_cells_by = "clusters_9",
           label_cell_groups = FALSE,
           label_leaves = TRUE, 
           label_branch_points = FALSE,
           alpha = 0.66) +
  scale_color_manual(values = rbow_pal) +
  labs(subtitle = x) + 
  lims(x = x_rng, y = y_rng) + 
  theme_nothing
})

p <- plot_grid(plotlist = plts, nrow = 1, ncol = 3)

save_plot(
    file.path(fig_dir, "supplement-7e-monocle_all_cells.pdf"),
    p,
    base_asp = 0.75,
    nrow = 1,
    ncol = 3
  )
p
```

```{r}
zoom_x <- c(-0.5, 2.25)
zoom_y <- c(4.75, 7.5)
plts <- lapply(levels(cds$treatment), function(x) {
  plot_cells(cds[, colData(cds)$treatment == x], 
           color_cells_by = "clusters_9",
           label_cell_groups = FALSE,
           label_leaves = TRUE, 
           label_branch_points = FALSE,
           alpha = 0.66, cell_size = 1) +
  scale_color_manual(values = rbow_pal) +
  labs(subtitle = x) + 
  lims(x = zoom_x, y = zoom_y) + 
  theme_nothing
})

p <- plot_grid(plotlist = plts, nrow = 1, ncol = 3)

save_plot(
    file.path(fig_dir, "supplement-7e-monocle_zoom_cells.pdf"),
    p,
    base_asp = 0.75,
    nrow = 1,
    ncol = 3
  )
p
```


# Summary files 

```{r}
library(Seurat)
sce <- readRDS(file.path(obj_dir, "20221230_scanvi_sce.rds"))
scanvi_avg_log2_expr <- assay(summarizeAssayByGroup(sce, 
                                                    sce$clusters_9,
                                                    statistics = "mean",
                                                    assay.type = "logcounts"))

so <- as.Seurat(sce)
so <- ScaleData(so, features = rownames(so))
Idents(so) <- "clusters_9"
scanvi_zscores <- AverageExpression(so,
                        slot = "scale.data", 
                        group.by = "clusters_9", 
                        return.seurat = FALSE)$originalexp

sce <- readRDS(file.path(obj_dir, "20221230_scvi_sce.rds"))
scvi_avg_log2_expr <- assay(summarizeAssayByGroup(sce,
                                                  sce$clusters_8,
                                                  statistics = "mean", 
                                                  assay.type = "logcounts"))

so <- as.Seurat(sce)
so <- Seurat::ScaleData(so, features = rownames(so))
Idents(so) <- "clusters_8"
scvi_zscores <- AverageExpression(so,
                             slot = "scale.data", 
                             group.by = "clusters_8", 
                             return.seurat = FALSE)$originalexp

out <- list(scanvi_avg_log2_expr = scanvi_avg_log2_expr, 
            scanvi_zscores = scanvi_zscores, 
            scvi_avg_log2_expr = scvi_avg_log2_expr,
            scvi_zscores = scvi_zscores)

out <- lapply(out, function(x){
  colnames(x) <- paste0("cluster_", colnames(x))
  x <- as.data.frame(cbind(gene = rownames(x), x))
  rownames(x) <- NULL
  class(x$gene) <- "Text"
  x
})
  
openxlsx::write.xlsx(out, 
                     "tables/average_cell_statistics.xlsx",
                     overwrite = TRUE)
    
```

cell type compositions

```{r}
cell_counts <- as.data.frame.matrix(table(sce$clusters_9, sce$study) )
fwrite(cbind(clusters = rownames(cell_counts), cell_counts),
       "tables/cells_per_cluster_per_study.csv",
       sep = ",",
       quote = FALSE, 
       col.names = TRUE, 
       row.names = FALSE)

tmp_sce <- sce[, sce$treatment == "Bleomycin" ]
cell_counts <- as.data.frame.matrix(table(tmp_sce$clusters_9,
                                          tmp_sce$strunz_et_al_timepoint) )
fwrite(cbind(clusters = rownames(cell_counts), cell_counts),
       "tables/bleo_cells_per_cluster_per_time_point.csv",
       sep = ",",
       quote = FALSE, 
       col.names = TRUE, 
       row.names = FALSE)
```

# session info

<details><summary>Show session info</summary>

```{r code}
sessionInfo()
```

</details>

