

### All cells together

```{r, echo = FALSE}
ps$all
```

### Split by study

```{r, echo = FALSE, fig.width = 9}
ps$by_study
```

### Old general cell types

```{r, echo = FALSE, fig.width = 9}
ps$all_old
```

### Cell type Markers {.tabset}


```{r, results ='asis', echo = FALSE}
for(i in seq_along(ps$genes)){
  cat('\n#### ', names(ps$genes)[i], '\n')
  print(ps$genes[i])
  cat('\n')
}
```

### CCA Markers {.tabset}

```{r, results ='asis', echo = FALSE}
for(i in seq_along(ps$cca_genes)){
  cat('\n#### ', names(ps$cca_genes)[i], '\n')
  print(ps$cca_genes[i])
  cat('\n')
}
```

### Activated Markers {.tabset}

```{r, results ='asis', echo = FALSE}
for(i in seq_along(ps$act_genes)){
  cat('\n#### ', names(ps$act_genes)[i], '\n')
  print(ps$act_genes[i])
  cat('\n')
}
```

### Clusters 

```{r}
cuts <- seq(4, 12)
ids <- paste0("clusters_", cuts)
if(!all(cuts %in% colnames(colData(sce)))){
  for(i in seq_along(cuts)){
    id <- ids[i]
    sce[[id]] <- as.character(igraph::cut_at(metadata(sce)$wlktrp, cuts[i]))
  }
}
```

Shown below is a "clustree". This is a diagram that shows how cells are related to 
each other at different clustering settings. Here I'm showing how the clustering changes
with increasing number of clusters (4-12). 

```{r, fig.height=9, fig.width=9}
library(clustree)
clustree(sce, prefix = "clusters_") +  scale_color_brewer(palette = "Set1")
```

### cluster comparisions to old integration {.tabset}

Showm below is the clustering results at different number of clusters. 

The UMAP shows the clusters, then there are 2 heatmaps which show the similarity 
bettwen old cell types or clusters and the new clustering. A jaccard index value of 1
means that the cells are identical between a pair of clusters, a value of 0 means no cells are shared.


```{r, results ='asis', echo = FALSE}
library(magrittr)
bpp <- BiocParallel::MulticoreParam(6)
dir.create(mkrs_dir, showWarnings = FALSE)

for(i in seq_along(ids)){
  cat('\n#### ', ids[i], '  \n')
  id <- ids[i]
  p <- plotUMAP(sce, 
         point_size = 0.5, 
         colour_by = id, 
         other_fields = "study") + 
     facet_wrap(~study)
  print(p)
  cat('  \n')
  
  cat("<br>\n\n<br>Relationship between old coarse cell types and new clustering.  \n")
  jacc.mat <- linkClustersMatrix(sce[[id]], sce$old_coarse_cell_types)
  rownames(jacc.mat) <- paste("New :", rownames(jacc.mat))
  colnames(jacc.mat) <- paste("Old :", colnames(jacc.mat))
  hm <- Heatmap(jacc.mat,
                name = "jaccard",
                col = viridis::viridis(100), 
                cluster_columns = FALSE,
                cluster_rows = FALSE)
  draw(hm)
  
  cat("<br>\n\n<br>Relationship between old clustering and new clustering  \n")
  
  jacc.mat <- linkClustersMatrix(sce[[id]], sce$old_clusters)
  rownames(jacc.mat) <- paste("New cluster:", rownames(jacc.mat))
  colnames(jacc.mat) <- paste("Old cluster:", colnames(jacc.mat))
  hm  <- Heatmap(jacc.mat,
        name = "jaccard",
        col = viridis::viridis(100), 
        cluster_columns = FALSE,
        cluster_rows = FALSE)
  draw(hm)
  
  cat("<br>\n\n<br>Shown below are the top 10 markers of each cell population in a table.  \n")

  if(!id %in% names(metadata(sce))){
    int_method <- metadata(sce)$integration
    mkrs <- scoreMarkers(sce, 
                        groups = sce[[id]],
                        block = sce$study,
                        BPPARAM = bpp) 
    mkrs <- lapply(mkrs ,function(x){
        x <- as.data.frame(x)  
        x <- cbind(gene = rownames(x), x)
        x <-  x[x$mean.AUC > 0.6, ]
        x[order(-x$mean.AUC, -x$mean.logFC.cohen), ]
      })
    
    metadata(sce)[[id]] <- mkrs
    mkrs %>%
      rbindlist(use.names = TRUE, idcol = "cluster") %>% 
      fwrite(file.path(mkrs_dir, paste0(id, "_markers.csv")), sep = ",")
    openxlsx::write.xlsx(mkrs, file.path(mkrs_dir, paste0(int_method, "_markers.xlsx")),
                         overwrite = TRUE)
  } else {
    mkrs <- metadata(sce)[[id]]
  }
  
  topx <- rbindlist(mkrs, use.names = TRUE, idcol = "group") %>% 
    .[, group := factor(group, levels = sort(unique(sce[[id]])))] %>% 
    .[, .SD[1:10], by = group] 
  
  res <- topx %>%
    .[, .(gene, group)] %>% 
    .[, id := 1:.N, by = group] %>% 
    dcast(id  ~ group, value.var = "gene") %>% 
    .[, !c("id")] %>% 
    as.data.frame()
  
  print(knitr::kable(res))
  
  cat('\n\n<!-- -->\n\n')
}
```

```{r}
saveRDS(sce, sce_fn)
```


