---
title: "Meta-analysis of murine lung regeneration studies: prelude"
author: "Kent Riemondy"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: kate
    fig_caption: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ComplexHeatmap)
library(SingleCellExperiment)
library(scran)
library(scuttle)
library(scater)
library(data.table)
library(here)
library(readxl)
library(openxlsx)
data_dir <- here("data")
formatted_data_dir <- "processed_data"
fig_dir <- "figs"

dir.create(fig_dir, showWarnings = FALSE)
dir.create(formatted_data_dir, showWarnings = FALSE)

obj_dir <- "objects"
dir.create(obj_dir, showWarnings = FALSE)
```

For each study, a study specific R script  (found in data/) will preprocess the data from public sources and save the dataset in a .rds file. 

## Gene lists

```{r}
studies <- c(
  "Kobayashi_et_al_ncb" = "41556_2020_542_MOESM2_ESM.xlsx",
  "Riemondy_et_al_JCI-insight" = "2.14.2019_Suppl_Table_5.xlsx",
  "Strunz_et_al_nc" = "41467_2020_17358_MOESM4_ESM.xlsx",
  "Wu_et_al_cell" = "Nan Tang Table S2.xlsx"
  )

studies <- setNames(file.path(data_dir, names(studies), studies), names(studies))

combined_markers <- list()
combined_tables <- list()
combined_expression <- list()
```

## Kobayashi_et_al_ncb


>Kobayashi, Y., Tata, A., Konkimalla, A. et al. Persistence of a regeneration-associated, transitional alveolar epithelial cell state in pulmonary fibrosis. Nat Cell Biol 22, 934–946 (2020). https://doi.org/10.1038/s41556-020-0542-8

Found a transitional cell state in epithelial lung organoids. 

```{r}
study_id <- "Kobayashi_et_al_ncb"
mkrs <- studies[[study_id]]

tbl <- read_excel(mkrs, skip = 1)[, 1:4] |> as.list()
names(tbl) <- gsub("...[0-9]$", "", names(tbl))
tbl <- lapply(seq_along(tbl), function(i){
  x <- tbl[[i]]
  data.frame(cell_type = names(tbl)[i],
             gene = x[!is.na(x)])
})
tbl <- do.call(rbind, tbl)

combined_markers[[study_id]] <- tbl[tbl$cell_type == "PATS", ]$gene
combined_tables[[study_id]] <- tbl

sce_fn <- file.path(data_dir, study_id, "sce.rds")

if(!file.exists(sce_fn)) source(file.path(data_dir, "process_ncb.R"))

sce <- readRDS(sce_fn)
avg_expr <- assay(aggregateAcrossCells(sce,
                                       ids = sce$cell_type,
                                       statistics = "mean",
                                       use.assay.type = "logcounts")) |>
  as.data.frame()

colnames(avg_expr) <- paste0(study_id, "::", colnames(avg_expr))
combined_expression[[study_id]] <- avg_expr
```

## Riemondy_et_al_JCI-insight

```{r}
study_id <- "Riemondy_et_al_JCI-insight"
mkrs <- studies[[study_id]]

tbl <- read_excel(mkrs, sheet = 2,
                  col_names = c(
                    "Gene",
                    "avg_logFC",
                    "pct1",
                    "pct2",
                    "pval",
                    "adj_p_val"
                  ), skip = 1) 
tbl <- tbl[tbl$adj_p_val < 0.05, ]
tbl <- tbl[order(tbl$adj_p_val), ] 
tbl <- as.data.frame(tbl)

combined_markers[[study_id]] <-tbl$Gene
combined_tables[[study_id]] <- tbl

sce_fn <- file.path(data_dir, study_id, "sce.rds")
if(!file.exists(sce_fn)) source(file.path(data_dir, "process_jci-insight.R"))
sce <- readRDS(sce_fn)

avg_expr <- assay(aggregateAcrossCells(sce,
                                       ids = sce$pretty_cell_labels,
                                       statistics = "mean",
                                       use.assay.type = "logcounts")) |>
  as.data.frame()

avg_expr <- avg_expr[, c("Naive AEC1",
                         "Naive AEC2",
                         "Injured AEC2: Early Transdifferentiating",
                         "Injured AEC2: Late Transdifferentiating", 
                         "Injured AEC2: Proliferating", 
                         "Other Injured AEC2"), drop = FALSE]
colnames(avg_expr) <- paste0(study_id, "::", colnames(avg_expr))
combined_expression[[study_id]] <- avg_expr
```

## Strunz_et_al_nc


>
Strunz M, Simon LM, Ansari M, et al. Alveolar regeneration through a Krt8+ transitional stem cell state that persists in human lung fibrosis. Nat Commun [Internet]. 2020;11:3559. Available from: http://dx.doi.org/10.1038/s41467-020-17358-3.

```{r}
study_id <- "Strunz_et_al_nc"
mkrs <- studies[[study_id]]

tbl <- read_excel(mkrs, sheet = 2)

to_keep <- tbl$`annotated cluster` ==	"Krt8+ ADI" & 
  tbl$p_val_adj < 0.05 & 
  tbl$avg_logFC > 0

tbl <- tbl[to_keep, ]
tbl <- tbl[order(tbl$p_val_adj), ] 
tbl <- as.data.frame(tbl)

combined_markers[[study_id]] <- tbl$gene
combined_tables[[study_id]] <- tbl


sce_fn <- file.path(data_dir, study_id, "sce.rds")
if(!file.exists(sce_fn)) source(file.path(data_dir, "process_nc.R"))

sce <- readRDS(file.path(data_dir, study_id, "sce.rds"))
avg_expr <- assay(aggregateAcrossCells(sce,
                                       ids = sce$cell_type,
                                       statistics = "mean",
                                       use.assay.type = "logcounts")) |>
  as.data.frame()

avg_expr <- avg_expr[,
                     c("Krt8+ ADI",
                       "AT1",
                       "AT2",
                       "AT2 activated",
                       "Mki67+ Proliferation")]
colnames(avg_expr) <- paste0(study_id, "::", colnames(avg_expr))
combined_expression[[study_id]] <- avg_expr
```

## Wu_et_al_cell

>Wu H, Yu Y, Huang H, et al. Progressive Pulmonary Fibrosis Is Caused by Elevated Mechanical Tension on Alveolar Stem Cells. Cell [Internet]. 2020;180:107–121.e17. Available from: http://dx.doi.org/10.1016/j.cell.2019.11.027.

```{r}
study_id <- "Wu_et_al_cell"
mkrs <- studies[[study_id]]

tbl <- read_excel(mkrs, skip = 1) 
tbl <- tbl[tbl$`fold change (subpopulation I AT2 cells/subpopulation II AT2 cells)` > 0, ]

combined_markers[[study_id]] <- tbl$`Gene name`
combined_tables[[study_id]] <- tbl

avg_expr_d7 <- read_excel(file.path(data_dir, study_id, "C7.average.exp.xlsx"))
colnames(avg_expr_d7)[1] <- "mouse_gene_name" 
colnames(avg_expr_d7)[2:3] <- paste0(study_id, "::", colnames(avg_expr_d7)[2:3], "_d7")

sce_fn <- file.path(data_dir, study_id, "sce.rds")
if(!file.exists(sce_fn)) source(file.path(data_dir, "process_cell.R"))

sce <- readRDS(sce_fn)
avg_expr <- assay(aggregateAcrossCells(sce,
                                       ids = sce$cell_type,
                                       statistics = "mean",
                                       use.assay.type = "logcounts")) |>
  as.data.frame()

colnames(avg_expr) <- paste0(study_id, "::", colnames(avg_expr))
combined_expression[[study_id]] <- avg_expr
```


## Save 

```{r}
saveRDS(combined_markers, file.path(formatted_data_dir, "gene_lists.rds"))
saveRDS(combined_tables, file.path(formatted_data_dir, "gene_tables.rds"))
saveRDS(combined_expression, file.path(formatted_data_dir, "avg_expression.rds"))
```

## session info

<details><summary>Show session info</summary>

```{r code}
sessionInfo()
```

</details>
